---
title: "Initial Gene Expression Analysis of RNA-Seq V2 RSEM TCGA Data for Selected Genes (both normal and tumor samples)"
author: "Amy Olex"
date: "Today's Date"
output: 
  html_document:
    toc: yes
    toc_depth: 3
---

Overview
----------

We are studying Genes associated with **INSERT PATHWAY AND CANCER** cancer in RNASeqV2 gene expression data from TCGA.  The RNASeq data was obtained via the Illumina HiSeq platform.  This report contains the analysis of all normal and tumor samples.

Methods
---------

- I parsed the RNASeqV2 raw_counts data for all patients of the selected cancer type using my tcgaRC.R class.
- Using methods in the tcgaRC class I parsed out selected clinical data of each sample, and classified each sample as tumor or normal.
- I used R to perform various normalizations of the RNASeq data prior to plotting. The specific normalization method used in this report is printed out below.
- I used all normal and tumor samples for this analysis.

```{r env_setup, echo=FALSE, message=FALSE, warning=FALSE}
######### Set up my environment ###########
library(DESeq2)
library(gplots)
library(ggplot2)
library(corrplot)
library(plyr)
library(reshape2)
library(grid)
library(gridExtra)

###Set new working directory
setwd("/Users/alolex/Desktop/CCTR/Data/KazuakiTakabe_S1P_0714_TCGA/TCGA Analyses/RNAseq_PAAD_all_111914/")

####Specify inputfile that has all normalized data
file <- "/Users/alolex/Desktop/CCTR/Data/KazuakiTakabe_S1P_0714_TCGA/TCGA Analyses/RNAseq_PAAD_all_111914/normalization/paad_111914_allSamplesNormalized_n183.RData"

###Define genes of interest with aliases
alias = data.frame(row.names=c( "EMR1.2015","TNF.7124","IL6.3569","STAT3.6774","NFKB1.4790","NFKB2.4791","RELA.5970","PECAM1.5175","LYVE1.10894","SPHK1.8877","SPHK2.56848","S1PR1.1901","S1PR2.9294","S1PR3.1903","S1PR4.8698","S1PR5.53637","ABCC1.4363","ABCG2.9429","SPNS2.124976","FLT1.2321","KDR.3791","FLT4.2324","TIE1.7075","TEK.7010","VEGFA.7422","VEGFC.7424","ANGPT2.285","SGPL1.8879","SPP1.6696", "ABCC11.85320"), alias=c("F4/80","TNF-alpha","IL-6","Stat3","NFkB1","NFkB2","p65","CD31","Lyve-1","SphK1","SphK2","S1PR1","S1PR2","S1PR3","S1PR4","S1PR5","ABCC1","ABCG2","Spns2","VEGFR1","VEGFR2","VEGFR3","Tie1","Tie2","VEGFA","VEGFC","Ang2","S1P-Lyase","SPP1", "ABCC11"))


```

```{r define_functions, echo=FALSE, message=FALSE, warning=FALSE}

my_cor_plot <- function(gene_data, plot_num, title_base){
  
  cor_data = cor(t(gene_data))
  
  corrplot(cor_data, method="square", type="full", mar=c(1,0,4,0), order="hclust", hclust.method="complete", tl.col="black", addrect=9, title=paste("Plot ", plot_num, " - Correlation Matrix\n", title_base, " (N=",dim(gene_data)[2],")\nTop 9 Hierarchical Clusters Shown (Complete Linkage)", sep=""), tl.cex=.7)
  
  return(plot_num+1)
}

sub_analysis <- function(gene_data, plot_num, title_base, sub_targets, colors, fact){

#transpose the gene matrix
gene_data_t = as.data.frame(t(gene_data))
gene_data_subset = as.data.frame(gene_data_t[sub_targets])

#Euclidean Distance UnScaled Heatmap
par(xpd=TRUE)
heatmap.2(as.matrix(t(gene_data_subset)),ColSideColors=as.character(colors), trace="none", scale="none", xlab="Patient Samples", ylab="Genes", main=paste("Plot ", plot_num, "\n", title_base, " (N=",dim(gene_data_subset)[1],")\nScale = none, Dist = Euclid", sep=""), labCol="", margins=c(2,10), cexRow=1.5)
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(fact), col=levels(colors), lty=1, lwd=5, cex=.5)

plot_num = plot_num +1

#Euclidean Distance Scaled Heatmap
par(xpd=TRUE)
heatmap.2(as.matrix(t(gene_data_subset)),ColSideColors=as.character(colors), trace="none", scale="row", xlab="Patient Samples", ylab="Genes", main=paste("Plot ", plot_num, "\n", title_base, " (N=",dim(gene_data_subset)[1],")\nScale = row, Dist = Euclid", sep=""), labCol="", margins=c(2,10), cexRow=1.5)
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(fact), col=levels(colors), lty=1, lwd=5, cex=.5)
plot_num = plot_num +1

#Correlation Distance UnScaled Heatmap
par(xpd=TRUE)
heatmap.2(as.matrix(t(gene_data_subset)),ColSideColors=as.character(color), trace="none", scale="none", xlab="Patient Samples", ylab="Genes", main=paste("Plot ", plot_num, "\n", title_base, " (N=",dim(gene_data_subset)[1],")\nScale = none, Dist = Corr", sep=""), labCol="", margins=c(2,10), distfun=function(x) as.dist((1-cor(t(x))/2)), cexRow=1.5)
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(fact), col=levels(colors), lty=1, lwd=5, cex=.5)
plot_num = plot_num +1

#Correlation Distance UnScaled Heatmap
par(xpd=TRUE)
heatmap.2(as.matrix(t(gene_data_subset)),ColSideColors=as.character(color), trace="none", scale="row", xlab="Patient Samples", ylab="Genes", main=paste("Plot ", plot_num, "\n", title_base, " (N=",dim(gene_data_subset)[1],")\nScale = row, Dist = Corr", sep=""), labCol="", margins=c(2,10), distfun=function(x) as.dist((1-cor(t(x))/2)), cexRow=1.5)
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(fact), col=levels(colors), lty=1, lwd=5, cex=.5)
plot_num = plot_num +1

return(plot_num)
}

my_factor_boxplot <- function(sorted_data, fact, fact_name, plot_num, colors){
  #add factor designations to sorted matrix
sorted_data_t = as.data.frame(t(sorted_data))
sorted_data_t[,fact_name] = fact

test.m <- melt(sorted_data_t)
print(ggplot(test.m, aes_string(x="variable", y="value", fill=fact_name)) + 
  geom_boxplot() + scale_fill_manual(values=colors) + 
  scale_x_discrete(labels=row.names(sorted_data)) + 
  theme(plot.margin=unit(c(6,6,5,11),"mm"), 
        axis.title.x = element_text(face="bold",size=16),
        axis.text.x = element_text(angle=90, vjust=0.5, size=12), 
        axis.title.y = element_text(vjust=2, face="bold", size=16),
        axis.text.y = element_text(vjust=.5, size=12), 
        legend.title = element_text(size=12, face="bold"), 
        legend.text = element_text(size = 10, face = "bold"), 
        legend.background = element_rect(fill=NA,size=.5, linetype="dotted"), 
        legend.justification = c(1.03,-0.06), 
        legend.position=c(1.03,-0.06),
        plot.title = element_text(lineheight=.8, face="bold")) + 
  ggtitle(paste("Plot ",plot_num," - RNASeqV2\nSimple Quantile Normalization of Selected Genes (N=",length(sorted_data),")\nSorted by mean across all samples." ,sep="")) + 
  xlab("Genes") + 
  ylab("Gene Expression (log2)") +
  scale_y_continuous(breaks=c(5,6,7,8,9,10,11,12,13,14,15)) )

plot_num = plot_num +1

return(plot_num)
}


```

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
############ Load Pre-Normalized Data and Drug Data #############
# Get the RNASeq pre-normalized data
load(file)
# variables now avaliable from RData file:
#"data", "rounded_data", "coldata", "dds", "vsd", "vsd_data", "logboxcox_data", "logshift_data", "qnorm_data", "voom_cyc_data", "voom_scale_data", "voom_none_data", "voom_quant_data"
#I'm interested in the coldata and the qnorm_data
#since I may want to generate a different report using a different normalization, I will make it generic and easily changable here.

#change the following line of code to use a different normalization method
ndata <- qnorm_data
nname <- "Simple Quantile"

#ndata <- as.data.frame(voom_none_data$E)
#nname <- "Voom (none)"

#ndata <- vsd_data
#nname <- "VST"

#initilize plot_num tracker
plot_num=1
```

###Target Genes

The target genes for this analysis are:

**INSERT LIST OF TARGET GOI HERE**

```{r specify_targets, echo=FALSE, message=FALSE, warning=FALSE}
# transpose the vsd_data matrix
ndata_t = as.data.frame(t(ndata))

#rename the genes to remove the pipe | in the name for easier parsing.
newnames = gsub("\\|", ".", colnames(ndata_t))
colnames(ndata_t) = newnames

#Extract out the genes of interest from vsd_data_t and transpose it again at the same time so the genes are in the rows again
ndata_gene = as.data.frame(t(ndata_t[,row.names(alias)]))

#Now merge this with the Alias object and get the alias names as the row names
ndata_gene2 = merge(alias, ndata_gene, by=0, all=TRUE)
row.names(ndata_gene2) = ndata_gene2$alias
ndata_gene2 = ndata_gene2[,-(1:2)]

```

###Data Normalization

```{r, echo=FALSE, message=FALSE, warning=FALSE, size=12}
paste("The data being used is from the file: ", file)
paste("The data normalization used is: ", nname)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plot(density(as.matrix(logshift_data)), main="Density of UnScaled-Normalized-Rounded RSEM Read Counts\nfor Multiple Normalization Methods", xlim=c(-10,20), ylim=c(0,.5),col="blue")
lines(density(as.matrix(qnorm_data)), col="red")
lines(density(as.matrix(vsd_data)), col="green")
lines(density(voom_quant_data$E), col="lightblue")
lines(density(voom_scale_data$E), col="pink")
lines(density(voom_cyc_data$E), col="lightgreen")
lines(density(voom_none_data$E), col="grey")
legend("topright", legend = c("Shifted Log", "Simple Qnorm", "VST","Voom (quantile)","Voom (scale)", "Voom (Cyclic Loess)", "Voom (none)"), fill=c("blue", "red", "green", "lightblue", "pink", "lightgreen", "grey"))

```

###Boxplots Before and After Normalization

For this analysis I chose to keep all tumor and normal samples together to see if there were any major gene expression differences between tumor and normal samples for our genes of interest.  Therefore, I normalized across all samples.


```{r normalization_boxplots1, echo=FALSE, message=FALSE, warning=FALSE}
### Box plot BEFORE Normalization
# I can't plot all samples, so take a sample of these to plot
color = revalue(coldata$tumor_status, c("tumor"="gold", "normal"="lightblue"))
my_fact=coldata$tumor_status
color_legend = revalue(levels(my_fact), c("tumor"="gold", "normal"="lightblue"))
fact_name="tumor_status"

#color = revalue(coldata$histological_type, c("Colon Adenocarcinoma"="red","Colon Mucinous Adenocarcinoma"="green","na"="white"))
#my_fact = coldata$histological_type

#color = revalue(coldata$gender, c("MALE"="blue","FEMALE"="pink"))
#my_fact = coldata$gender

#color = revalue(coldata$DFS, c("free"="green", "recurrence"="red", "na"="white"))
#my_fact = coldata$DFS

#color = revalue(coldata$OS, c("alive"="green", "dead"="red", "na"="white"))
#my_fact = coldata$OS

#color = revalue(coldata$anatomic_site, c("Ascending Colon"="red","Cecum"="green","Descending Colon"="blue","Hepatic Flexure"="gold","na"="white","Rectosigmoid Junction"="lightblue","Sigmoid Colon"="lightgreen","Splenic Flexure"="black","Transverse Colon"="orange"))
#my_fact = coldata$anatomic_site

#color = revalue(coldata$race, c("AMERICAN INDIAN OR ALASKA NATIVE"="red", "ASIAN"="lightblue","BLACK OR AFRICAN AMERICAN"="black",       "na"="white", "WHITE"="green" ))
#my_fact = coldata$race


sample = seq(from=1, to=length(rounded_data), by=3)
boxplot(log(as.data.frame(rounded_data[sample]), base=2), col=as.character(color[sample]), main=paste("Plot ",plot_num,"\nRNASeqV2 Log2 Transformed Un-Normalized Data\nAll Samples, N=", length(rounded_data), ", Subset Plotted=", length(sample)), xlab="Samples", ylab="raw counts")

plot_num = plot_num+1

### Box plot AFTER Normalization
boxplot(as.data.frame(ndata[sample]), col=as.character(color[sample]), main=paste("Plot ",plot_num,"\nRNASeqV2 ",nname," Normalized Data\nAll Samples, N=", length(ndata), ", Subset Plotted=", length(sample)), xlab="Samples", ylab="normalized counts")

plot_num = plot_num+1
```

Box Plots for Genes of Interest
--------------------------------

```{r, echo=FALSE, message=FALSE, error=FALSE, fig.width=9, fig.height=7}
#sort the data based on mean across all samples
means = data.frame(row.names=row.names(ndata_gene2), Means = rowMeans(ndata_gene2))
ndata_gene3 = merge(means, ndata_gene2, by=0, all=TRUE)
row.names(ndata_gene3) = ndata_gene3$Row.names
ndata_gene3 = ndata_gene3[,-1]
ndata_gene3=ndata_gene3[order(ndata_gene3$Means),]

par(mar=c(8,5,5,3), mgp=c(3,1,0))
boxplot(t(ndata_gene3[,-1]), col="grey", main=paste("Plot ",plot_num," - RNASeqV2\nSimple Quantile Normalization of Selected Genes (N=",length(ndata),")\nSorted by mean across all samples." ,sep=""), las=2, xlab="", ylab="Gene Expression (log2)", xaxt="n")
axis(side=1, at=1:length(row.names(ndata_gene3)), labels=row.names(ndata_gene3), las=2)
mtext("Genes", side=1, line=6)

plot_num = plot_num+1
```

```{r,eval=FALSE, echo=FALSE, message=FALSE, error=FALSE}
# Now lets plot using ggplot2
test.m2 <- melt(t(ndata_gene3[,-1]))

# I can't figure out how to make all the boxes gold!
ggplot(test.m2, aes(x=Var2, y=value)) + 
  geom_boxplot() + 
  scale_x_discrete(labels=row.names(ndata_gene3)) + 
  theme(plot.margin=unit(c(6,6,5,11),"mm"), 
        axis.title.x = element_text(face="bold",size=20),
        axis.text.x = element_text(angle=90, vjust=0.5, size=16), 
        axis.title.y = element_text(vjust=2, face="bold", size=20),
        axis.text.y = element_text(vjust=.5, size=16), 
        legend.title = element_text(size=16, face="bold"), 
        legend.text = element_text(size = 16, face = "bold"), 
        legend.background = element_rect(fill="gray90", size=.5, linetype="dotted"), 
        plot.title = element_text(lineheight=.8, face="bold")) + 
  ggtitle(paste("Plot ",plot_num," - RNASeqV2\nSimple Quantile Normalization of Selected Genes (N=",length(ndata),")\nSorted by mean across all samples." ,sep="")) + 
  xlab("Genes") + 
  ylab("Gene Expression (log2)")

plot_num = plot_num+1
```

```{r, echo=FALSE, message=FALSE, error=FALSE, fig.width=9, fig.height=7}

plot_num <- my_factor_boxplot(ndata_gene3[,-1], my_fact, fact_name , plot_num, color_legend)

```

###Genes that are significantly different between tumor and normal samples at a p-value <=0.01

```{r, echo=FALSE, error=FALSE}
pval_cutoff = 0.01

tmp <- lapply(as.data.frame(t(ndata_gene3[,-1])), function(x) t.test(x ~ my_fact)$p.val)
str <- paste(names(as.data.frame(tmp[tmp<=pval_cutoff])),format(as.data.frame(tmp[tmp<=pval_cutoff]), digits=4), sep=":")
#str2 <- paste(str, collapse=",")
#print(str2[])
tmp.sig <- format(as.data.frame(t(as.data.frame(tmp[tmp<=pval_cutoff]))), digits=4)
names(tmp.sig) <- "P-value"
grid.table(tmp.sig)
```


Results
--------

###Sample Sizes

*Number of samples in full dataset (includes tumor and normal)*
```{r, echo=FALSE}
length(ndata)
```

*Number of samples that are from tumors*
```{r, echo=FALSE}
count(coldata$tumor_status[coldata$tumor_status=="tumor"])[1,2]
```

*Number of samples that are normal*
```{r, echo=FALSE}
count(coldata$tumor_status[coldata$tumor_status=="normal"])[1,2]
```


###Gene Expression Heatmap

This first heatmap shows the **un-scaled** gene expression for all patient samples (x-axis) for our genes of interest (y-axis).  Genes and samples were clusered using Euclidean distance, and the colors shown are unscaled.  Euclidean distance clusters on magnitude of gene expression, thus, it makes sense that the lowly expressed genes (red) are clustered together, and the highly expressed genes (yellow) are clustered together.   

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# create the heatmaps
par(xpd=TRUE)
heatmap.2(as.matrix(ndata_gene2),ColSideColors=as.character(color), trace="none", scale="none", xlab="Patient Samples", ylab="Genes", main=paste("Plot ",plot_num,"\nAll Samples (N=",dim(ndata_gene2)[2],")\nScale = none, Dist = Euclid", sep=""), labCol="", margins=c(2,10))
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(my_fact), col=levels(color), lty=1, lwd=5, cex=.5)

plot_num = plot_num+1
```

This next plot is the same as the previous except that I **scaled** the data for each gene across all patient samples.  Thus, gene expression is now represented as a z-score.  Essentially, scaling gives a similar z-score to the highest value in a lowly expressed gene and the highest value in a highly expressed gene.  However, using Euclidean distance to cluster and looking at scaled values doesn't tell us much.  When using Euclidean to cluster I think the unscaled values tell us more about the data.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
par(xpd=TRUE)
heatmap.2(as.matrix(ndata_gene2),ColSideColors=as.character(color), trace="none", scale="row", xlab="Patient Samples", ylab="Genes", main=paste("Plot ",plot_num,"\nAll Samples (N=",dim(ndata_gene2)[2],")\nScale = row, Dist = Euclid", sep=""), labCol="", margins=c(2,10))
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(my_fact), col=levels(color), lty=1, lwd=5, cex=.5)
plot_num = plot_num+1
```

Instead of clustering on magnitude of expression (Euclidean distance) I have now clustered on correlation of expression patterns across all patients (for the gene clustering) and across all genes (for the patient clustering).  The first plot here is the **un-scaled** data.  You can see some highly and lowly expressed genes are now clustered together, but from looking at the unscaled data it is not clear why.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

par(xpd=TRUE)
heatmap.2(as.matrix(ndata_gene2),ColSideColors=as.character(color), trace="none", scale="none", xlab="Patient Samples", ylab="Genes", main=paste("Plot ",plot_num,"\nAll Samples (N=",dim(ndata_gene2)[2],")\nScale = none, Dist = Corr", sep=""), labCol="", margins=c(2,10), distfun=function(x) as.dist((1-cor(t(x))/2)) )
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(my_fact), col=levels(color), lty=1, lwd=5, cex=.5)
plot_num = plot_num+1

```

For the remainder of the plots I will show both the **scaled** and **un-scaled** data, however, I think from our discussions the unscaled data for euclidean distance gives us the most information when clustering on magnitude of expression, and the scaled data for correlation gives us the most information when clustering on pattern of expression.

**Note, regardless of scaled or un-scaled heatmaps, the clustering is done BEFORE the scaling of values.**


```{r, echo=FALSE, message=FALSE, warning=FALSE}

par(xpd=TRUE)
heatmap.2(as.matrix(ndata_gene2),ColSideColors=as.character(color), trace="none", scale="row", xlab="Patient Samples", ylab="Genes", main=paste("Plot ",plot_num,"\nAll Samples (N=",dim(ndata_gene2)[2],")\nScale = row, Dist = Corr", sep=""), labCol="", margins=c(2,10), distfun=function(x) as.dist((1-cor(t(x))/2)) )
# create a legend
legend("topright", inset=c(-.01,-.1),legend=levels(my_fact), col=levels(color), lty=1, lwd=5, cex=.5)
plot_num = plot_num+1

```


###Correlation Heatmap

The following plot shows the correlation of gene profiles across all samples for each gene of interest. Blue is positively correlated and red is negatively correlated.  Results were hierarchically clustered before being plotted, and the top 7 clusters are shown with black squares around them.    

If you compare this plot with **Plot 3** and **Plot 4** you will see the groupings are the same because they are clustered using correlation. 

```{r first_plots_c, echo=FALSE, message=FALSE, warning=FALSE}
## Now create a correlation matrix for all samples
plot_num <- my_cor_plot(ndata_gene2, plot_num, "All Patient Samples")

```


Subgroup Analyses (CHANGE TO BE SPECIFIC TO CORRELTION HEATMAP FOR EACH ANALYSIS)
--------------------------------------------

From the correlation heatmap in **Plot 7** we see there are several subgroups of genes that have strong positive or negative correlation to each other.  I took some of these groups and created heatmaps of just those genes to see if the patient samples would cluster differently.  

###Analysis of (SUBGROUP OF GENES LISTED HERE)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# define the targets
sub_targets = c("Spns2", "Ang2", "S1PR5", "ABCC1", "VEGFA")
#transpose the gene matrix
plot_num <- sub_analysis(gene_data=ndata_gene2, plot_num=plot_num, title_base="All Patient Samples", sub_targets=sub_targets, colors=color, fact=my_fact)
```
